---
title: "Assignment 3 - 20200891"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r}
load("meat.Rdata")
load("oliveoil.Rdata")
```

# Task 1 - 1 - Construct a vector of wavelength values from 400 to 2498 steps in steps of 2

```{r}
Wavelength_Vector <- seq(from = 400, to = 2498, by = 2)
```


# Task 1 - 2 - Plot the spectrum of the first observation(row) against the wavelength values as a curve

```{r}
meat_wave_plot <- as.matrix(meatdata[,-1])
View(meat_wave_plot)
plot(meat_wave_plot[1,], main = "Spectrum of Wavelength Values 1st Observation",
     xlab = "Wavelength Spectrums", ylab = "Reflectance Values", type = "l")

```


# Task 1 - 3 - Calculate the median, 1st Quartile and 3rd Quartile of each column of the data 

```{r}

summ <- apply(meatdata[,-1],2,quantile,prob=c(0.25,0.50,0.75))
summ[,seq(from=1,to=1050,by=100)]

```


# Task 1 - 4 - Produce a plot that show the median, 1st Quartile and 3rd Quartile for each wavelength

```{r}

matplot(Wavelength_Vector,t(summ),type="l",xlab="Wavelength",ylab="Reflectance")
legend("topleft",legend=c("Q1","Median","Q3"),lty=1:3,col=1:3)

```


# Task 1 - 5 - Calculate the median value of each each column of the data for each meat type and plot these

```{r}
medianmat <- aggregate(meatdata[,-1],by=list(meatdata[,1]),median)
matplot(Wavelength_Vector,t(medianmat[,-1]),type="l",xlab="Wavelength",ylab="Reflectance")
legend("topleft",legend=names(table(meatdata[,1])),col=1:5,lty=1:5,cex=0.7)

```

# Task 1 - 6 - Create a new factor variable that combines the Chicken and Turkey sample into Poultry

```{r}
type <- meatdata[,1]
newtype <- as.character(type)
newtype[newtype=="Chicken"] <- "Poultry"
newtype[newtype=="Turkey"] <- "Poultry"
newtype <- factor(newtype)

```

# Task 1 - 7 - Re-do calculation in 1-6 with the new factor variable as the meat type

```{r}
medianmat <- aggregate(meatdata[,-1],by=list(newtype),median)
matplot(Wavelength_Vector,t(medianmat[,-1]),type="l",xlab="Wavelength",ylab="Reflectance")
legend("topleft",legend=names(table(newtype)),col=1:5,lty=1:5,cex=0.7)

```

# Task 2-1 - Write a function that reads in a data frame of the same format as the meat data (with a possibly different number of rows) that does the following:

# 2a - It computes the median, 1st Quartile and 3rd Quartile for each wavelength for each sample type

```{r}

spectrum_summary <- function(dat)
{
sample_type <- dat[,1]
spectra <- dat[,-1]
type_names <- names(table(sample_type))
n_types <- length(type_names)
summary_list <- list()
plot(Wavelength_Vector,spectra[1,],type="n",xlab="Wavelength",ylab="Reflectance",ylim=range(spectra))
for (i in 1:n_types)
{
type <- type_names[i]
summ <- apply(spectra[sample_type==type,],2,quantile,prob=c(0.25,0.50,0.75))
rownames(summ) <- c("Q1","Median","Q3")
4
matplot(Wavelength_Vector,t(summ),type="l",col=rep(i,3),lty=c(2,1,2),add=TRUE)
summary_list[[type]] <- summ
}
abline(v=700,col="darkgray")
legend("topleft",type_names,col=1:n_types,lty=rep(1,n_types),cex=0.5)
invisible(summary_list)
}

```

# 2b - It produces plot of the median, 1st Quartile and 3rd Quartile for each wavelength for each sample type

```{r}

meat_summary <- spectrum_summary(meatdata)

```

# 2c - A vertical line is added to the plot at 700nm; this is the boundary between visible and near infrared light


# 2d - The summary statistics computed in 2a are only reported if the output of the function is assigned to an object

```{r}
meat_summary[["Beef"]][,1:4]
```

# Task 2-2 - Use function to examine the data stored in oliveoil.Rdata.

```{r}

olive_summary <- spectrum_summary(oliveoil)

```

# Task 3 - Creativity

```{r}
#one-way ANOVA to see which wavelengths have the greatest evidence of difference between types of olive oil.


anova_fun <- function(x,l)
{
fit <- aov(x~l)
p_val <- anova(fit)$`Pr(>F)`[1]
return(p_val)
}
pvec <- apply(oliveoil[,-1],2,anova_fun,l=oliveoil[,1])
plot(Wavelength_Vector,pvec,type="l")
abline(h=0.05,col="red")
abline(v=700,col="darkgray")

#that most of the wavelengths have p-values less than 0.05, apart from some regions. Interesting, many of the wavelengths in the visible range have p-values greater than 0.05; which indicates that they don’t have significant evidence of difference.

```
